---
title: 'MLE: Lab 3'
author: "Shahryar Minhas"
date: "January 29, 2014"
header-includes:
   - \usepackage{multirow}
   - \usepackage{dcolumn}
output: 
  pdf_document:
    fig_caption: yes
---

# Setup

```{r, message=FALSE, warning=FALSE}
# Start with a clean workspace
rm(list=ls())

# Function to load packages
loadPkg=function(toLoad){
  for(lib in toLoad){
	if(! lib %in% installed.packages()[,1])
	  { install.packages(lib, repos='http://cran.rstudio.com/') }
	suppressMessages( library(lib, character.only=TRUE) ) }
}

# Load libraries
packs=c("ggplot2", 'lmtest', 'car', 'sandwich')
loadPkg(packs)

# Set a theme for gg
theme_set(theme_bw())

# Functions that I use frequently
char = function(x){ as.character(x) }
num = function(x){ as.numeric(char(x)) }

# Relevant paths
labPath='~/Dropbox/Duke/Spring 2015/PS 733/lab3'
```

# Muller & Seligson Regression

```{r, results='hide'}
# Load data
msrepPath=paste0(labPath, "/Msrepl87.asc")
msrep = read.table(msrepPath, header=TRUE)

# Create silly logged version of DV
msrep$deaths75ln = log(msrep$deaths75+1)

# Create logs of other things
msrep$deaths70ln = log(msrep$deaths70+1)
msrep$sanctions75ln = log(msrep$sanctions75+1)
msrep$sanctions70ln = log(msrep$sanctions70+1)
msrep$energypcln = log(msrep$energypc+1)
```

## Functional Form

### Omitted Variable Bias

```{r}
# How to assess effects of misspecification from OVB 
# Bivariate case
cor(msrepRegData[,c('deaths75ln', 'upper20', 'sanctions75ln')])

coeftest( lm(deaths75ln ~ upper20 + sanctions75ln, data=msrepRegData) )

# If we excluded sanctions75ln from the model what would you expect of the
# coefficient on upper20
coeftest( lm(deaths75ln ~ upper20, data=msrepRegData) )

# If we excluded upper20 from the model what would you expect of the
# coefficient on sanctions75ln
coeftest( lm(deaths75ln ~ sanctions75ln, data=msrepRegData) )

# More complicated model
# Harder to assess direction of bias in these cases
ivs=c('upper20', 'energypcln', 'intensep', 
      'sanctions70ln', 'sanctions75ln', 'deaths70ln')
msrepRegData=na.omit(msrep[,c('deaths75ln', ivs)])
olsForm1=formula(paste0('deaths75ln ~ ', paste(ivs, collapse=' + ')))
olsForm2=formula(paste0('deaths75ln ~ ', paste(ivs[c(2:length(ivs))], collapse=' + ')))
mod1 = lm(olsForm1, data=msrepRegData)
mod2 = lm(olsForm2, data=msrepRegData)

# View model results
summary(mod1)
summary(mod2)
```

### F-test

```{r}
# How would we do a partial F-test
ssr1=sum(resid(mod1)^2)
ssr2=sum(resid(mod2)^2)
chgReg=ncol(model.matrix(mod1)) - ncol(model.matrix(mod2))
# F statistic
Fstat=((ssr1-ssr2)/chgReg)/(ssr1/df.residual(mod1))
1-pf(abs(Fstat), chgReg, df.residual(mod1))

# Function from base stats
anova(mod1, mod2)
```

### Likelihood Ratio Test

```{r}
# This runs a likelihood ratio test
# Inputs are lm model objects, where the first is the
# full model and the second the restricted
likeRatTest <- function(full,restr){
  n=length(resid(full))
  ssrF=sum(resid(full)^2)
  ssrR=sum(resid(restr)^2)
  val=n*log(ssrR/ssrF)
  chgReg=ncol(model.matrix(full)) - ncol(model.matrix(restr))
  pval = 1-pchisq(val,chgReg)
  output=round(c(val, pval),5)
  names(output)=c('Likelihood Ratio', 'P-value')
  return(output)  
  }
likeRatTest(mod1,mod2)


```

### Partial Residual Plots

```{r}
# How to test for non-linear effects in parameters
# Can use a the crPlots function from the car library
# to generate partial residual plots
crPlots(mod1)

# Incorporating a non-linear effect

msrepRegData$upper20sq = msrepRegData$upper20^2
olsForm3=formula('deaths75ln ~ upper20 + upper20sq + energypcln + intensep + 
                 sanctions70ln + sanctions75ln + deaths70ln')

olsForm3=formula('deaths75ln ~ poly(upper20,2) + energypcln + intensep + 
                 sanctions70ln + sanctions75ln + deaths70ln')
mod3=lm(olsForm3, data=msrepRegData)

# Examine coefficient results
coeftest(mod3)

# How to evaluate whether to keep non-linear term?
anova(mod1, mod3)
lrtest(mod1, mod3)
```

### Non-Constant Coefficients

The Chow test is a statistical and econometric test of whether the coefficients in two linear regressions on different data sets are equal. The Chow test was invented by economist Gregory Chow in 1960. In econometrics, the Chow test is most commonly used in time series analysis to test for the presence of a structural break.

The Chow test is used to see if it makes sense to run two separate regressions on two mutually exclusive subsets of your data (divided by a break point) by comparing the results of the two “unrestricted” regressions versus the “restricted” regression that pools all the data together.

The procedure is as follows:
* Run a “restricted” regression on all your data (pooled).
* Divide your sample into to groups, determined by your breakpoint (e.g. a point in time, or a variable value).
* Run an “unrestricted” regression on each of your subsamples.  You will run two “unrestricted” regressions with a single breakpoint.


```{r, results='hide', include=FALSE}
# Computing the chow statistic
# First convert data into two lets say random subsamples
set.seed(6886)
par(mfrow=c(1,1))
data(faithful)
of = faithful
plot(of, col="blue",main="Eruptions of Old Faithful")

## Run three regressions (1 restricted, 2 unrestricted)
r.reg = lm(waiting ~ eruptions, data = of)
ur.reg1 = lm(waiting ~ eruptions, data = of[of$eruptions > 3.25,])
ur.reg2 = lm(waiting ~ eruptions, data = of[of$eruptions <=3.25,])
## review the regression results
coeftest(r.reg)
coeftest(ur.reg1)
coeftest(ur.reg2)
 
## Calculate sum of squared residuals for each regression
SSR = NULL
SSR$r = r.reg$residuals^2
SSR$ur1 = ur.reg1$residuals^2
SSR$ur2 = ur.reg2$residuals^2
 
## K is the number of regressors in our model
K = r.reg$rank
 
## Computing the Chow test statistic (F-test)
numerator = ( sum(SSR$r) - (sum(SSR$ur1) + sum(SSR$ur2)) ) / K
denominator = (sum(SSR$ur1) + sum(SSR$ur2)) / (nrow(of) - 2*K)
chow = numerator / denominator
chow
 
## Calculate P-value
1-pf(chow, K, (nrow(of) - 2*K))

## Plot the results
plot(of,main="Eruptions of Old Faithful")
# restricted model
abline(r.reg, col = "red",lwd = 2, lty = "dashed")
# unrestricted model 1
segments(0, ur.reg2$coefficients[1], 3.25,
ur.reg2$coefficients[1]+3.25*ur.reg2$coefficients[2], col= 'blue')
# unrestricted model 2
segments(3.25, ur.reg1$coefficients[1]+3.25*ur.reg1$coefficients[2],
5.2, ur.reg1$coefficients[1]+5.2*ur.reg1$coefficients[2], col= 'blue')

## Sort the data
sort.of = of[order(of$eruptions) , ]
sort.of = cbind(index(sort.of),sort.of)
 
## Identify the row number of our breakpoint
brk = max(sort.of[,1][sort.of$eruptions<=3.25])
 
## Using the CRAN package 'strucchange'
library(strucchange)
sctest(waiting ~ eruptions, type = "Chow", point = brk, data = sort.of)

of$dummy = as.numeric(of$eruptions >= 3.25)
summary(lm(eruptions ~ waiting + I(dummy*waiting) + dummy, data = of))
```

## Heteroskedasticity

```{r}
# How to calculate the Breusch-Pagan test statistic
residMod1=resid(mod1)^2
bpForm=formula(paste0('residMod1 ~', paste(ivs, collapse=' + ')))
bpMod=lm(bpForm, data=msrepRegData)
bpStat=summary(bpMod)$r.squared*nrow(msrepRegData)
1-pchisq(bpStat, df=length(ivs))

# Breusch-Pagan test: using bptest from lmtest library
bptest(mod1)
```

```{r}
# Residuals vs fitted
par(mfrow=c(1,2))
plot(mod1, 1)
plot(predict(mod1), resid(mod1))

# Scale-location
stdResid=(resid(mod1)-mean(resid(mod1)))/sd(resid(mod1))
plot(mod1, 3)
plot(predict(mod1), sqrt(abs(stdResid)))
```

#### Dealing with heteroskedasticity: Robust standard errors

```{r, results='hide'}
# Can use the sandwich package in combination with lmtest to help
# us calculate robust standard errors
coeftest(mod1, vcov=vcovHC(mod1, type='HC1'))
coeftest(mod1)
```